```{r setup, include=FALSE,message=FALSE,warning=FALSE}
# OPTIONS -----------------------------------------------
knitr::opts_chunk$set(echo = TRUE, 
                      warning=FALSE, 
                      message = FALSE)

# PACKAGES-----------------------------------------------
# Tutorial packages
library(vembedr)
library(skimr)
library(yarrr)
library(RColorBrewer)
library(GGally) 
library(tidyverse)
library(plotly)
library(readxl)
library(rvest)
library(biscale)
library(tidycensus)
library(cowplot)
library(units)
library(olsrr)

data("HousesNY", package = "Stat2Data")
HousesNY$Beds <- as.numeric(HousesNY$Beds)



```

# Summarizing Data {#T7_summary}

Before analyzing a dataset, it's essential to understand its structure, size, and data types. This tutorial will guide you through:

1.  Viewing the dataset.
2.  Checking its dimensions, column names, and structure.
3.  Understanding data types and modifying them if necessary.
4.  Preparing for summary statistics.

### Suggested libraries

```{r}
library(dplyr)
library(skimr)
library(GGally)
```

### How to edit

The tutorial is based on a small dataset called `test_data` with these four columns,

-   `Make` (categorical)
-   `MPG` (numerical)
-   `Cylinders` (categorical)
-   `Transmission` (categorical)

So to edit for your code, all you need to do is replace `test_data` with the name of your spreadsheet/data-frame, and to replace the variable you are looking at with your column name e,g,

`test_data$MPG` goes to `WhateverYourDataIsCalled$ColumnName`

Here is the full dataset.

```{r, include=FALSE}
# Create a simple dataset with numerical and categorical variables
test_data <- data.frame(
  Make = factor(c("Toyota", "Ford", "Honda", "Chevy", "Nissan", "BMW","Toyota", "Ford", "Honda","Toyota")),
  MPG = c(30, 25, 27, 22, 29, 24, 29,23,25,25),
  Cylinders = factor(c(4, 6, 4, 8, 4, 6,8,4,5,4)),
  Transmission = c("Auto", "Manual", "Auto", "Auto", "Manual", "Manual","Auto","Auto","Auto","Manual")
)
```

```{r}
print(test_data)
```

<br>

## How to view data

To explore the dataset interactively, you can:

-   **Open it in the RStudio Data Viewer**:\
    Run `View(test_data)` in the console.\
    *Note: `View()` should not be placed inside a code chunk because it only works interactively in RStudio and does not produce console output. It often breaks the knitting process*

-   **See a tabular preview in the environment pane**:\
    Click on `test_data` in the environment tab.

-   **Print the entire dataset** (not recommended for large datasets):

```{r}
# this will only show the first page
test_data
```

-   **View only the first six rows**:

```{r}
head(test_data)
```

<br>


### Print column names 

To see what columns exist:

```{r}
names(test_data)
```



## How to summarise data

<br>

### Number of Rows/Columns

To find out how many rows and columns there are:

```{r}
nrow(test_data)  # Number of rows
ncol(test_data)  # Number of columns
dim(test_data)   # Dimensions (rows, columns)
```

<br>



### Checking data types

Each column in a dataset has a specific data type (e.g., numeric, character, factor). "Factor" is the R jargon for variables that have a fixed number of unique values (e.g., categories/groups/families).

We can check the data types using:

```{r}
sapply(test_data, class)
```

<br>

If categorical variables are not stored as factors, we can explicitly define them, where levels is the ORDER I want them to be in.

```{r}
test_data$Cylinders    <- factor(test_data$Cylinders, levels = c(4, 6, 8))
test_data$Transmission <- factor(test_data$Transmission, levels = c("Auto", "Manual"))
```

This ensures:

-   Consistent ordering of categorical levels.
-   Avoiding unintended sorting in plots or models.

Similarly, if data "should" be numeric, but isn't for some reason, we can convert it using as.numeric() etc.

<br><br>




## Summary commands

There are many commands!  You don't need all of these - choose your favourite :)

### str()


To get a compact summary of the dataset, `str()` provides an overview, including the number of observations, variables, and data types.

```{r}
str(test_data)
```

<br>

### dplyr::glimpse()

Another alternative is `glimpse()` (from the `dplyr` package, so you have to have run the library code chunk first):

```{r}
glimpse(test_data)
```
<br>

### summary()


We can use the `summary()` function to get a quick overview of numerical and factor variables.

```{r}
summary(test_data)
```

<br>


### skimr:skim()`

The `skim` command from the `skimr` package provides an easy and readable summary of the dataset

```{r}
skim(test_data)
```

<br>


### Frequency tables


We can use `table()` or `dplyr::count()` to summarize categorical variables. e.g. count how many rows there are in different categories. This is especially useful for categorical data.

The easiest way to do this is using the `table` command. 

```{r}
table(test_data$Transmission)
```

Shows there are three rows with automatic cars and three with manual, 

We can also make two way tables

```{r}
table(test_data$Transmission, test_data$Cylinders)
```

So there are 2 cars that are automatic with four cylinders. 

We can also use the `count` function in the dplyr package.

```{r}
test_data %>% count(Transmission)
```





### Calculate your own

Alternatively, we can calculate specific summary statistics manually using base R and `dplyr`.

```{r}

mean  (test_data$MPG, na.rm=TRUE)  # Mean MPG, ignoring missing values
median(test_data$MPG, na.rm=TRUE)  # Median MPG, ignoring missing values
sd    (test_data$MPG, na.rm=TRUE)      # Standard deviation, ignoring missing values
range (test_data$MPG, na.rm=TRUE)   # Minimum and maximum, ignoring missing values

```


### Grouped Summary Statistics

Finally,  we can also compute statistics such as the mean, maximum, and minimum values for different groups.

Using `group_by()` and `summarise()` from dplyr() , we can compute statistics for each species:

```{r}
test_data %>%
  dplyr::group_by(Transmission) %>%
  dplyr::summarise(
    Mean_MPG = mean(MPG),
    Median_MPG = median(MPG),
    Count = n()
  )
```

<br>



