```{r setup, include=FALSE,message=FALSE,warning=FALSE}
# OPTIONS -----------------------------------------------
knitr::opts_chunk$set(echo = TRUE, 
                      warning=FALSE, 
                      message = FALSE)

# PACKAGES-----------------------------------------------
# Tutorial packages
library(vembedr)
library(skimr)
library(yarrr)
library(RColorBrewer)
library(GGally) 
library(tidyverse)
library(plotly)
library(readxl)
library(rvest)
library(biscale)
library(tidycensus)
library(cowplot)
library(units)
library(olsrr)

data("pirates", package = "yarrr")

```



# Transformations {#S.Tutorial_Transformations}

We will talk in the class (or read here! https://online.stat.psu.edu/stat501/lesson/9 ) about *why* to apply transformations.  Here I will simply show you how.

We simply make a new column and use maths to fill in the value.

For example. Here is some data.

```{r,message=FALSE}
# read the data
data <- read.csv("neither.csv")
head(data)
```

To make a column with $\frac{1}{x}$ as the predictor, I would use the $ to create a new column a and on the other side of the arrow, take the fraction.

```{r,message=FALSE}
# Note I am trying to avoid spaces  or special characters in my data
data$transform_1overx <- 1/data$x
```


To make a column with $x^2$ as the predictor, we do

```{r,message=FALSE}
# Note I am trying to avoid spaces  or special characters in my data
data$transform_xsquare <- data$x^2
```

To make a column with $\log(x)$ as the predictor, we do (this is the NATURAL LOG)

```{r,message=FALSE}
# Note I am trying to avoid spaces  or special characters in my data
data$transform_lnx <- log(data$x)
```

To make a column with $\log_{10}(x)$ as the predictor, we do 

```{r,message=FALSE}
# Note I am trying to avoid spaces  or special characters in my data
data$transform_logx <- log(data$x,base=10)
```

To make a column with $\sqrt{x}$ as the predictor, we do 

```{r,message=FALSE}
# Note I am trying to avoid spaces  or special characters in my data
data$transform_sqrtx <- sqrt(data$x)
```

And let's see the output

```{r}
head(data)
```

We could now perform our linear regression with any of these as our predictor.  For example


```{r}
newmodel <- lm(y~transform_sqrtx,data=data)
plot(data$y~data$transform_sqrtx,xlab="Sqrt(x) UNITS",ylab="y UNITS")
```

You don't *have* to make a new column in advance, you can do the maths within the regression itself:

```{r}
newmodel2 <- lm(y~sqrt(x),data=data)

# For Polynomial regression you can use the poly command e.g. X+X^2 

newmodel3 <- lm(y ~ poly(x,2),data=data)

```

```{r}
summary(newmodel3)
```

```{r}
equatiomatic::extract_eq(newmodel3)
```

